From clusters to queries: exploiting uncertainty in the modularity landscape of complex networks

### 2018-05-02 article/article.tex
sign

The index \textit{coverage} is directly proportional to the number of starting random partitions.
A full coverage index could be considered as every locally optimal partition.
Given that there is no free lunch and it is impossible to know every solution, one can only ensure a full coverage index through an exhaustive search over the $2^m$ possible starting cut sets.
Consequently, in the approach taken here is to use a large but not exhaustive subset of the possible solutions using 10,000 solutions for the large networks studied in this paper.

@@

\subsection{Statistically significant query sets}
\label{sec:query_sign}
Naturally, not every label set will be of high quality or represented within the community structure of a network.
This is established further in section \ref{sec:go_labels} which tests gene ontology terms (GO) on protein interaction networks.
In this case, all the labels relate to the real function of the vertices, however, GO is a hierarchical system and many terms are extremely general.
Furthermore, even for specific terms the network datasets will often not be well represented.
A common approach with community detection and clustering (used for example in \cite{dekkers2013}) is to detect communities and find clusters containing significant GO terms.

In this work, we would wish to use the community partition space, described above, to find highly related sets of vertices.
For this we take the distribution, from Equation \ref{eq:mu_score} for a given query set.
For a query set to be relevant it should be significantly more related than one would expect from a set of nodes selected at random within a network.

Formally, given the partition space $\mathcal{P}$ and a query set $S$ we consider the distribution of quality scores as $\mu_i(S - i)$, for all $i$ in $S$.
In order for $S$ to be a high quality query set, its value of internal vertices should be significantly higher than for external vertices.
Formally, we assert that the distribution $\mu_i(S - i)$ should reject the null hypothesis that the mean is not significantly greater than $\mu_j(S)$ where $j \in S - V$, the set of vertices excluding those in the query set.
For the purposes of this work we use the non-parametric Mann–Whitney $U$ test and reject the null hypothesis where $p < 0.001$.
In section \ref{sec:go_labels} this is used to exclude less relevant GO terms.

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.23\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/lfr_binary_mo_overlap_auc_1000.eps}
        \caption{1000 nodes}
    \end{subfigure}
    \begin{subfigure}[b]{0.23\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/lfr_binary_mo_overlap_auc_5000.eps}
        \caption{5000 nodes}
    \end{subfigure}
    \caption{Non-overlapping community varying seed nodes on LFR networks with 1000 and 5000 nodes.
     Data points represent mean auc scores for all communities on 100 sampled networks at varying mixing coefficients.
     Error bars represent standard deviations.}
     \label{fig:auc_no_overlap}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.23\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/lfr_binary_overlap_auc_1000.eps}
        \caption{1000 nodes}
    \end{subfigure}
    \begin{subfigure}[b]{0.23\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/lfr_binary_overlap_auc_5000.eps}
        \caption{5000 nodes}
    \end{subfigure}
    \caption{Overlapping community varying seed nodes on LFR networks with 1000 and 5000 nodes.
     Data points represent mean auc scores for all communities on 100 sampled networks at varying overlapping fraction of nodes.
     Error bars represent standard deviations.}
     \label{fig:auc_overlap}
\end{figure}

\subsection{Real networks}

@@

In order to test the performance of the semi-supervised classification on real-world data we present our findings on example networks with known metadata communities.
All datasets taken use the largest single connected component sub-graph.

\subsubsection{Dataset descriptions}
\label{sec:real_network_descs}
The real networks used are listed as follows:
\begin{itemize}
 \item \textbf{EU emails dataset (EU emails)} This anonymised dataset is taken from the SNAP database \cite{snap} and contains $986$ nodes and $16,687$ edges representing emails between individuals.
 The metadata community labels represent different departments within the organisation.
 In total there are 42 communities, 39 of which contain at least 3 nodes.

@@

 \item \textbf{\textit{Arabidopsis thaliana} protein-protein interaction network (Arabidopsis PPI)} \cite{ArabidopsisConsortium2011}

 \item \textbf{\textit{Escherichia coli} protein-protein interaction network (E.coli PPI)}

\end{itemize}

@@

    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/rocs/arabidopsis_ppi.png}
        \caption{Arabidopsis PPI protein complexes}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/rocs/ecoli_ppi.png}
        \caption{E coli PPI protein complexes}
    \end{subfigure}
    \caption{ROC scores for the high quality labels for real networks described in Section \ref{sec:real_network_descs}. 
    Results represent the averages over all reciever operating characteristic scores from all samples of label sets with the cross validation procedure described in \ref{sec:cross_validation}.
    Shaded areas indicate standard deviations.}
    \label{fig:quality_label_rocs}
\end{figure*}
% 
% 
\subsubsection{High quality labels}
In this section we observe the performance of our method on the real networks described in Section \ref{sec:real_network_descs} with high quality, gold standard, curated labels.
For the EU Emails dataset, the communities relate to real departments of individuals.
For the protein interaction datasets we used manually curated interaction sets.
In the case of the Yeast this was taken from experimentally validated complexes  \cite{pu2008up}.
For the \textit{E coli} dataset we used manually curated interacting proteins from \cite{}.
For the Arabidopsis dataset, the complex sources were more limited.
Consequently, the IntAct database was used with all protein complexes listed containing less than 10 proteins in total \cite{}.

The results with varying number of seed nodes, for the cross validation procedure described in Section \ref{sec:cross_validation} are shown in Figure \ref{fig:quality_label_rocs}.
These results represent ROC and AUC statistics

\subsubsection{Gene ontology data}
\label{sec:go_labels}
In this section we analyse the performance of our method on gene ontology data for the protein interaction networks in this study.
Unlike the the curated protein complexes and known departmental memberships in Section \ref{sec:real_network_labels}, gene ontology cannot be considered fully relevant in all cases.
Specific terms may have no relevance in the context of protein-protein interactions, and many terms are extremely general.
Consequently, we cannot expect the labels to perform well in this context.
Therefore, we use the query set significance testing outlined in Section \ref{sec:query_sign}.
The data used in this study contains all non-redundant gene ontology terms available for proteins contained within a network from AmiGO \cite{carbon2008amigo}.
The purpose of this test is not to find biologically relevant information but, rather, to show how the method presented here can be tuned to evaluate poor quality label sets.

@@

\section{Conclusions}
This paper has presented a novel approach to semi-supervised community detection utilising a consensus of high scoring partitions computed with the popular modularity maximisation approach.
Previously the glassy search space of this optimisation algorithm has been seen as a major limitation of the approach.
However, in this work we consider each locally optimal partition to be information regarding the true multi-class labels that are likely present in real networks.
As the algorithm is trivial to run in a distributed manner, this approach can be applied to relatively large graphs.
Further research can be conducted into this approach with alternative quality functions to modularity that do not suffer the same limitations in terms of a resolution limit.

However, this approach requires the community landscape to contain many local maxima, a property likely shared by many real-world, heterogeneous networks.
Similarly, the method presented here requires both some labelled data and the labels to be represented well within the topology of the underlying network.
The approach presented here differs from other ensemble approaches in that the objective is to provide a probabilistic framework for label classification.
This approach performs extremely well on both synthetically generated networks and real-world ground truth communities with relatively small sets of labels.
However, in a practical application this would still require selecting some form of arbitrary cut off for defining \textit{crisp} communities.

This also work presents a number of interesting potential future avenues for research, such as observing how query sets change in time dynamic or multi-scale networks.

>>>>

The index \textit{coverage} is directly proportional to the number of starting random partitions.
A full coverage index could be considered as every locally optimal partition.
Given that there is no free lunch and it is impossible to know every solution, one can only ensure a full coverage index through an exhaustive search over the $2^m$ possible starting cut sets.
Consequently, in the approach taken here is to use a large but not exhaustive subset of the possible solutions using 2000 solutions for the networks studied in this paper.

@@

% \subsection{Statistically significant query sets}
% \label{sec:query_sign}
% Naturally, not every label set will be of high quality or represented within the community structure of a network.
% This is established further in section \ref{sec:go_labels} which tests gene ontology terms (GO) on protein interaction networks.
% In this case, all the labels relate to the real function of the vertices, however, GO is a hierarchical system and many terms are extremely general.
% Furthermore, even for specific terms the network datasets will often not be well represented.
% A common approach with community detection and clustering (used for example in \cite{dekkers2013}) is to detect communities and find clusters containing significant GO terms.
% 
% In this work, we would wish to use the community partition space, described above, to find highly related sets of vertices.
% For this we take the distribution, from Equation \ref{eq:mu_score} for a given query set.
% For a query set to be relevant it should be significantly more related than one would expect from a set of nodes selected at random within a network.
% 
% Formally, given the partition space $\mathcal{P}$ and a query set $S$ we consider the distribution of quality scores as $\mu_i(S - i)$, for all $i$ in $S$.
% In order for $S$ to be a high quality query set, its value of internal vertices should be significantly higher than for external vertices.
% Formally, we assert that the distribution $\mu_i(S - i)$ should reject the null hypothesis that the mean is not significantly greater than $\mu_j(S)$ where $j \in S - V$, the set of vertices excluding those in the query set.
% For the purposes of this work we use the non-parametric Mann–Whitney $U$ test and reject the null hypothesis where $p < 0.001$.
% In section \ref{sec:go_labels} this is used to exclude less relevant GO terms.

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/lfr_binary_mo_overlap_auc_1000.eps}
        \caption{1000 nodes}
    \end{subfigure}

    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/lfr_binary_mo_overlap_auc_5000.eps}
        \caption{5000 nodes}
    \end{subfigure}
    \caption{Non-overlapping community varying seed nodes on LFR networks with 1000 and 5000 nodes.
     Data points represent mean auc scores for all communities on 100 sampled networks at varying mixing coefficients.
     Error bars represent standard deviations.}
     \label{fig:auc_no_overlap}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/lfr_binary_overlap_auc_1000.eps}
        \caption{1000 nodes}
    \end{subfigure}

    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/lfr_binary_overlap_auc_5000.eps}
        \caption{5000 nodes}
    \end{subfigure}
    \caption{Overlapping community varying seed nodes on LFR networks with 1000 and 5000 nodes.
     Data points represent mean auc scores for all communities on 100 sampled networks at varying overlapping fraction of nodes.
     Error bars represent standard deviations.}
     \label{fig:auc_overlap}
\end{figure}

\subsection{Real networks}
\label{sec:real_networks}

@@

In order to test the performance of the semi-supervised classification on real-world data we present our findings on example networks with known metadata communities.
All datasets taken use the largest single connected component sub-graph.
The real networks used are listed as follows:
\begin{itemize}
 \item \textbf{EU emails dataset (EU emails)} This anonymised dataset is taken from the SNAP database \cite{snap} and contains $986$ nodes and $16,687$ edges representing emails between individuals.
 The metadata community labels represent different departments within the organisation.
 In total there are 42 communities, 39 of which contain at least 3 nodes.

@@

 \item \textbf{\textit{Arabidopsis thaliana} protein-protein interaction network (Arabidopsis PPI)} \cite{ArabidopsisConsortium2011}
  For the Arabidopsis dataset, the complex sources were more limited.
  Consequently, the IntAct database was used with all protein complexes listed containing less than 10 proteins in total \cite{}.

  \item \textbf{\textit{Escherichia coli} protein-protein interaction network (E.coli PPI) \cite{}}
  For the \textit{E coli} dataset we used manually curated interacting proteins from \cite{}.
\end{itemize}

The results with varying number of seed nodes, for the cross validation procedure described in Section \ref{sec:cross_validation} are shown in Figure \ref{fig:quality_label_rocs}.
These results represent ROC curves and mean AUC statistics.
For all the datasets in this study binary community classification is high, with an average AUC score above 0.88, even when using only a single seed node.
In the case of the \textit{Arabidopsis thaliana} network the protein complexes tested are significantly smaller than for other networks and so a comparison of seed sizes is not possible.
However, in the other networks studied using 3 or more seed nodes appears to improve results, though for 15 seed nodes the results are not significantly better than with 7 seeds.

@@

    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/rocs/arabidopsis_ppi.png}
        \caption{Arabidopsis PPI protein complexes}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{images/rocs/ecoli_ppi.png}
        \caption{E coli PPI protein complexes}
    \end{subfigure}
    \caption{ROC scores for the high quality labels for real networks described in Section \ref{sec:real_networks}. 
    Results represent the averages over all reciever operating characteristic scores from all samples of label sets with the cross validation procedure described in \ref{sec:cross_validation}.
    Shaded areas indicate standard deviations.}
    \label{fig:quality_label_rocs}
\end{figure*}
.

@@

\section{Conclusions}
This paper has presented a novel approach to semi-supervised community detection utilising a consensus of high scoring partitions computed with the popular modularity maximisation approach.
Previously the glassy search space of this optimisation algorithm has been seen as a major limitation.
However, in this work we consider each locally optimal partition to be information regarding the true multi-class labels that are likely present in real networks.
The approach presented here differs from other ensemble approaches in that the objective is to provide a probabilistic framework for label classification.
Performance is extremely strong on both synthetically generated networks and real-world ground truth communities with relatively small sets of labels.

However, this approach requires the community landscape to contain many local maxima, a property likely shared by many real-world, heterogeneous networks.
Similarly, the method presented here requires both some labelled data and the labels to be represented well within the topology of the underlying network.

This also work presents a number of interesting potential future avenues for research, such as observing how query sets change in time dynamic or multi-scale networks.
Furthermore, as the algorithm is trivial to run in a distributed manner, this approach could be applied to larger graphs than those studied in this paper.
Further research should also be conducted into how this approach could be applied to other partition quality functions.
